Robustly decorrelating errors with mixed quantum gates

### 2019-12-13 decorrelation/mixed_unitaries.tex
Minor updates

In general an error will neither be fully coherent nor Pauli-stochastic and so it is useful to define what we mean by the coherence of arbitrary maps\cite{CarignanDugas2019}.
 One way to measure this is with a quantity called the unitarity\cite{Wallman2015}:
\begin{align}
u(\mathsf{E}) = \frac{d}{d-1}\int d\psi ||\mathbf{n}[\mathsf{E}(\psi)] - \mathbf{n}[\mathsf{E}(\identmat_d/2)||^2
\end{align}
where $\mathbf{n}$ is the length of the Bloch vector, and the integral is over all $d$-dimensional pure states using the Haar measure. Yet another way of quantitavely describing how unitary a map is with protocols that assess how errors accumulate on different input states\cite{1907.05950}. In this work, we will only discuss the relative unitarity of channels, and therefore are content if the unitarity of our error map decreases by any of these metrics.

@@

If the diamond distance error rates of the component gates are bounded by $\alpha$, then Campbell shows that this first-order cancellation of the error generators can ensure that the diamond distance error rate of the MQG is bounded by $\alpha^2$. For gate compilation errors, the error channels are dominated by unitary approximation error, so the error generators are Hamiltonians. Hamiltonian generators can have positive or negative coefficients, so it is possible that the origin may lie in their convex hull. If, however, the error map contains stochastic components, then the condition of Eq.~\eqref{eq:cancel} might be impossible to satisfy. Such errors always have strictly positive probabilities, so the origin can \emph{never} lie in their convex hull. This can be rectified by restricting the sum in Eq.~\eqref{eq:cancel} to only the Hamiltonian component of the generators, or by replacing the exact condition with a minimization, as we discuss in Sec.~\ref{sec:mixed_unitary_processes}.

@@

While generator-exact MQGs can provide a guaranteed suppression of the diamond distance, their effective error channels are unlikely to be a purely stochastic. Second- and higher-order terms in \eqref{eq:generators_and_effective_error} can easily contribute to lingering coherent errors that impact the efficient simulability of the channel. In order to definitively eliminate these coherent errors, we could instead seek weights, $\weight^P_i$, that annihilate the off-diagonal entries of the effective error map. The resulting diagonal error map will be Pauli-stochastic. For a single qubit gate, it takes the form,
% \begin{equation}
% 	\errmat_{eff} = 
% 	\left(\begin{array}{@{}ccccc@{}}
% 		\cdot & \cdot & \cdot & \cdot \\ 
%   		\cline{1-1}\cline{3-4}
%     	\multicolumn{1}{|c}{0} & \multicolumn{1}{|c}{\cdot} & \multicolumn{1}{|c}{0} & \multicolumn{1}{c|}{0} \\
%     	\cline{2-2}\cline{3-3}		
% 		\multicolumn{1}{|c}{0} & 0 & \multicolumn{1}{|c}{\cdot} & \multicolumn{1}{|c|}{0}\\
% 		\cline{3-3}\cline{4-4}	
% 		\multicolumn{1}{|c}{0} & 0 & 0 & \multicolumn{1}{|c}{\cdot}\\
% 		\cline{1-3}
% 	\end{array} 	
% 	\right).
% \end{equation}
\begin{equation}
	\errmat_{\rm{eff}} = \sum_i \weight_i^P \errmat_i = 
	\left(\begin{array}{@{}ccccc@{}}
		1 & 0 & 0 & 0 \\ 
    	0 &  p_x & 0 & 0 \\
		0 & 0 &  p_y & 0\\
		0 & 0 & 0 &  p_z\\
	\end{array} 	
	\right),
\end{equation}
where $p_x, p_y$ and $p_z$ are the rates of Pauli $X$, $Y$, and $Z$ errors, respectively. We refer to such channels as \emph{Pauli-exact} MQGs, and the same geometric argument we used in constructing generator-exact MQGs is again applicable here. The off-diagonal elements of the error map form a vector space, so we need only check that the origin is in the convex hull of the vectors of off-diagonal elements for each of the component operations, $\errmat_i$. If the error map contains non-unital components or correlated Pauli-stochastic errors (such as $X+Y$), then this strict geometric condition will not hold and no Pauli-exact MQG will exist. However, MQGs that approximate the Pauli-exact condition can be found via the convex programing techniques discussed in Sec.~\ref{sub:off_diagonals}.

@@

The remainder of this section will assume such a capability for generating a large ensemble of gate implementations, and will focus instead on the second step: constructing a distribution over that ensemble that reduces the coherence of the effective error channel. We will begin by discussing the two broad optimization targets introduced above: i) generator-exact MQGs and ii) Pauli-exact MQGs. We then propose a set of secondary targets that can improve the performance by i) incorporating robustness to drift, ii) targeting low-error-rate solutions, and iii) reducing the number of native unitaries that contribute to the effective channel.

@@

% ==============================================================================
% Section: Robustness
% ==============================================================================
\subsection{Secondary Objectives}
\noindent 
Often, the cost of generating distinct gate implementations is quite small, so large ensembles of them can be computed rather quickly. It is often the case that one will have available many more distinct gate implementations than there are parameters in the relevant vector space (see Fig.~\ref{fig:vectorspace}). This can lead to the the minimization problems discussed in the previous section being massively under-constrained, yielding a large continuous family of exact solutions. Secondary objectives provide a means for choosing among this family of exact MQGs those with increased performance in a desired area. In this section we present explicit convex programs for such secondary objectives, including adding robustness to drift or model uncertainty, reducing the effective gate error, and minimizing the number of constituent gates utilized by the MQG. We choose to take generator-exact MQGs as our starting point, though the following sections apply equally well to Pauli-exact MQGs with only trivial modifications.

@@

% ==============================================================================
% Section: Diamond Distance Inequality
% ==============================================================================
\subsection{Proof of diamond distance inequality}
\label{sub:diamond_distance_inequality}
Here we prove the claim of \eqref{eq:diamond_ineq} that:
\begin{equation}
	\dnorm\!\left(\error_{\rm eff}\right) \le \sum \weight_i \, \dnorm(\error_i).
\end{equation}
The effective error channel for a mixed quantum gate is $\error_{\rm{eff}} = \sum \weight_i\,\error_i$, where $\error_i$ are the error channels for the component gates. The diamond distance to the identity of the effective error channel is:
\begin{align}
	\dnorm \left(\error_{\rm eff}\right)
		&= \frac{1}{2} \sup_\rho \vert \vert (\ident_d\otimes \ident_d)(\rho) 
										  - (\error_{\rm eff} \otimes \ident_d)(\rho) \vert\vert_1\\
		% &= \frac{1}{2} \sup_\rho \vert \vert (\ident_d\otimes \ident_d)(\rho) 
										  % - (\sum_i \weight_i\,\error_i \otimes \ident_d)(\rho) \vert\vert_1 \\
		&= \frac{1}{2} \sup_\rho \vert \vert \sum \weight_i\,((\ident_d
										  - \error_i) \otimes \ident_d)(\rho) \vert\vert_1
\end{align}
For qubits, the space of density matrices is compact, so the supremum is achievable. Call a state that achieves the supremum $\rho^*$. Then 
\begin{align}
	\dnorm \left(\error_{\rm eff}\right)
		&= \frac{1}{2} \vert \vert \sum \weight_i\,((\ident_d
										  - \error_i) \otimes \ident_d)(\rho^*) \vert\vert_1 \\
		&= \frac{1}{2} \vert \vert \sum \weight_i\, \rho^*_i \vert\vert_1,
\end{align}
where we have defined $\rho^*_i = ((\ident_d - \error_i) \otimes \ident_d)(\rho^*)$. The nuclear norm above is equal to the sum of the singular values of $\sum \weight_i \rho_i$. Using the Ky Fan singular value inequality \cite{fan1951maximum} , we have 
\begin{align}
	\dnorm \left(\error_{\rm eff}\right)
		&\le \frac{1}{2} \sum_i \weight_i \vert \vert \rho^*_i \vert\vert_1 \\
		&\le \sum_i \weight_i\, \dnorm(\error_i)
\end{align}
The second inequality above follows because $\rho^*$ defines an explicit lower bound for the diamond distance for each of the component error maps.

>>>>

% In general an error will neither be fully coherent nor Pauli-stochastic and so it is useful to define what we mean by the coherence of arbitrary maps\cite{CarignanDugas2019}.
%  One way to measure this is with a quantity called the unitarity \cite{Wallman2015}:
% \begin{align}
% u(\mathsf{E}) = \frac{d}{d-1}\int d\psi ||\mathbf{n}[\mathsf{E}(\psi)] - \mathbf{n}[\mathsf{E}(\identmat_d/2)||^2
% \end{align}
% where $\mathbf{n}$ is the length of the Bloch vector, and the integral is over all $d$-dimensional pure states using the Haar measure. Yet another way of quantitavely describing how unitary a map is with protocols that assess how errors accumulate on different input states\cite{1907.05950}. In this work, we will only discuss the relative unitarity of channels, and therefore are content if the unitarity of our error map decreases by any of these metrics.

@@

If the diamond distance error rates of the component gates are bounded by $\alpha$, then Campbell shows that this first-order cancellation of the error generators can ensure that the diamond distance error rate of the MQG is bounded by $\alpha^2$. For gate compilation errors, the error channels are dominated by unitary approximation error, so the error generators are Hamiltonians. Hamiltonian generators can have positive or negative coefficients, so it is possible that the origin may lie in their convex hull. If, however, the error map contains stochastic components, then the condition of Eq.~\eqref{eq:cancel} might be impossible to satisfy. Such errors always have strictly positive probabilities, so the origin can \emph{never} lie in their convex hull. This formalism can be adapted to such cases by restricting the sum in Eq.~\eqref{eq:cancel} to only the Hamiltonian component of the generators, or by replacing the exact condition with a minimization, as we discuss in Sec.~\ref{sec:mixed_unitary_processes}.

@@

While generator-exact MQGs can provide a guaranteed suppression of the diamond distance, their effective error channels are unlikely to be a purely stochastic. Second- and higher-order terms in \eqref{eq:generators_and_effective_error} can easily contribute to lingering coherent errors that impact the efficient simulability of the channel. In order to definitively eliminate these coherent errors, we could instead seek weights, $\weight^P_i$, that annihilate the off-diagonal entries of the effective error map. The resulting diagonal error map will be Pauli-stochastic. For a single qubit gate, it takes the form,
% \begin{equation}
% 	\errmat_{eff} = 
% 	\left(\begin{array}{@{}ccccc@{}}
% 		\cdot & \cdot & \cdot & \cdot \\ 
%   		\cline{1-1}\cline{3-4}
%     	\multicolumn{1}{|c}{0} & \multicolumn{1}{|c}{\cdot} & \multicolumn{1}{|c}{0} & \multicolumn{1}{c|}{0} \\
%     	\cline{2-2}\cline{3-3}		
% 		\multicolumn{1}{|c}{0} & 0 & \multicolumn{1}{|c}{\cdot} & \multicolumn{1}{|c|}{0}\\
% 		\cline{3-3}\cline{4-4}	
% 		\multicolumn{1}{|c}{0} & 0 & 0 & \multicolumn{1}{|c}{\cdot}\\
% 		\cline{1-3}
% 	\end{array} 	
% 	\right).
% \end{equation}
\begin{equation}
	\errmat_{\rm{eff}} = \sum_i \weight_i^P \errmat_i = 
	\left(\begin{array}{@{}ccccc@{}}
		1 & 0 & 0 & 0 \\ 
    	0 &  p_x & 0 & 0 \\
		0 & 0 &  p_y & 0\\
		0 & 0 & 0 &  p_z\\
	\end{array} 	
	\right),
\end{equation}
where $p_x, p_y$ and $p_z$ are the rates of Pauli $X$, $Y$, and $Z$ errors, respectively. We refer to such channels as \emph{Pauli-exact} MQGs, and the same geometric argument we used in constructing generator-exact MQGs is again applicable here. The off-diagonal elements of the error map form a vector space, so we need only check that the origin is in the convex hull of the vectors of off-diagonal elements for each of the component operations, $\errmat_i$. If the error map contains non-unital components or correlated Pauli-stochastic errors (such as $X+Y$), then this strict geometric condition may not hold and no Pauli-exact MQG would exist. However, MQGs that approximate the Pauli-exact condition can be found via the convex programing techniques discussed in Sec.~\ref{sub:off_diagonals}.

@@

The remainder of this section will assume such a capability for generating a large ensemble of gate implementations, and will focus instead on the second step: constructing a distribution over that ensemble that reduces the coherence of the effective error channel. We will begin by discussing the two broad optimization targets introduced above: i) generator-exact MQGs and ii) Pauli-exact MQGs. We then propose a set of secondary optimization targets that can improve the performance by i) incorporating robustness to drift, ii) targeting low-error-rate solutions, and iii) reducing the number of native unitaries that contribute to the effective channel.

@@

% ==============================================================================
% Section: Robustness
% ==============================================================================
\subsection{Secondary Objectives}
\noindent 
Often, the cost of generating distinct gate implementations is quite small, so large ensembles of them can be computed rather quickly. One may then have available many more distinct gate implementations than there are parameters in the relevant vector space (see Fig.~\ref{fig:vectorspace}). This can lead to the the minimization problems discussed in the previous section being massively under-constrained, yielding a large continuous family of exact solutions. Secondary objectives provide a means for choosing among this family of exact MQGs those with increased performance in a desired area. In this section we present explicit convex programs for such secondary objectives, including adding robustness to drift or model uncertainty, reducing the effective gate error, and minimizing the number of constituent gates utilized by the MQG. We choose to take generator-exact MQGs as our starting point, though the following sections apply equally well to Pauli-exact MQGs with only trivial modifications.

@@

% ==============================================================================
% Section: Diamond Distance Inequality
% ==============================================================================
\subsection{Proof of diamond distance inequality}
\label{sub:diamond_distance_inequality}
Here we prove the claim of Eq.~\eqref{eq:diamond_ineq} that:
\begin{equation}
	\dnorm\!\left(\error_{\rm eff}\right) \le \sum \weight_i \, \dnorm(\error_i).
\end{equation}
The effective error channel for a mixed quantum gate is $\error_{\rm{eff}} = \sum \weight_i\,\error_i$, where $\error_i$ are the error channels for the component gates. The diamond distance to the identity of the effective error channel is:
\begin{align}
	\dnorm \left(\error_{\rm eff}\right)
		&= \frac{1}{2} \sup_\rho \vert \vert (\ident_d\otimes \ident_d)(\rho) 
										  - (\error_{\rm eff} \otimes \ident_d)(\rho) \vert\vert_1\\
		% &= \frac{1}{2} \sup_\rho \vert \vert (\ident_d\otimes \ident_d)(\rho) 
										  % - (\sum_i \weight_i\,\error_i \otimes \ident_d)(\rho) \vert\vert_1 \\
		&= \frac{1}{2} \sup_\rho \vert \vert \sum \weight_i\,((\ident_d
										  - \error_i) \otimes \ident_d)(\rho) \vert\vert_1
\end{align}
For qubits, the space of density matrices is compact, so the supremum is achievable. Call a state that achieves the supremum $\rho^*$. Then 
\begin{align}
	\dnorm \left(\error_{\rm eff}\right)
		&= \frac{1}{2} \vert \vert \sum \weight_i\,((\ident_d
										  - \error_i) \otimes \ident_d)(\rho^*) \vert\vert_1 \\
		&= \frac{1}{2} \vert \vert \sum \weight_i\, \rho^*_i \vert\vert_1,
\end{align}
where we have defined $\rho^*_i = ((\ident_d - \error_i) \otimes \ident_d)(\rho^*)$. The nuclear norm above is equal to the sum of the singular values of $\sum \weight_i \rho_i$. Using the Ky Fan singular value inequality \cite{fan1951maximum} , we have 
\begin{align}
	\dnorm \left(\error_{\rm eff}\right)
		&\le \frac{1}{2} \sum_i \weight_i \vert \vert \rho^*_i \vert\vert_1 \\
		&\le \sum_i \weight_i\, \dnorm(\error_i)
\end{align}
The second inequality above follows because $\rho^*$ defines an explicit lower bound for the diamond distance for each of the component error maps.
